Treatment heterogeneity in comparative effectiveness of teriparatide vs bisphosphonates: multi-database cohort study

### 2021-03-01 submission/manuscript.rmd
add more references

```{r, echo=FALSE, message=FALSE}
library(tidyverse)
library(pander)
library(knitr)

@@

- **IBM MarketScan Medicare Supplemental Database (MDCR)** is a claims database representing retirees in the United States with primary or Medicare supplemental coverage
- **Optum De-Identified Clinformatics® Data Mart Database – Date of Death (DOD)** is a claims database including members of private health insurance.

@@

Our cohorts included female new users of teriparatide or oral bisphosphonates above the age of 65 (ref 7,8 LEGEND-HTN). Patients were considered to be new-users if they received no anti-osteoporosis drugs (raloxifene, bazedoxifene, denosumab, abaloparatide, romosozumab) 365 days prior to treatment initiation with teriparatide or an oral bisphosphonate. We required patients to have at least one year of observation period before first exposure and an established osteoporosis diagnosis, defined as any recorded fracture of the hip, wrist, spine or shoulder/humerus in their recorded medical history.

@@

##     Statistical analyses
We carried out two sets of analyses. First, we derived overall treatment effect estimates of teriparatide compared to oral bisphosphonates regarding the three outcomes of interest. To account for potential measured confounding we developed separate propensity score models within each database using the same predefined set of measured covariates. Covariates were included in the propensity score model using regularized logistic regression (##REFERENCE Tian-Schuemie simulation). We estimated treatment effects using Cox proportional hazards models with treatment as the sole covariate fitted in the 1-4 propensity score-matched subset of the considered patient population. Results from different databases were summarized using random effects meta-analysis.

For the second set of analyses, we used the standardized framework for risk-based assessment of treatment effect heterogeneity, that follows the directives of the PATH statement (##REFERENCE framework). Following data extraction, the framework was applied in two steps. First, we derived individualized risk predictions for the three efficacy outcomes. We built the prediction models using LASSO logistic regression on the propensity score matched (1 to 4) subpopulation of the pooled treatment arms, aiming to remove any effect of measured confounding that may cause differential fit of our models across treatment arms. We considered the same large set of candidate covariates for the development of our prediction models.

For each outcome we used the derived prediction models to divide the population under study in 4 equally-sized subgroups, based on the quarters of the predicted risk distribution. Within each of these risk-based subgroups we developed a new propensity score model. Our analyses were performed on the propensity score matched (1 to 4) subset of the risk subgroup subset. We used a set of XXX negative control outcomes, to calibrate our results for unmeasured confounding (##REFEFRENCE). We derived relative effect estimates using Cox proportional hazards models only with treatment as a predictor. Absolute effect estimates were calculated based on the difference of the Kaplan-Meier estimates, on day 730 after treatment initiation.

@@

When assessing heterogeneity of treatment effect for hip fracture, we estimated calibrated hazard ratios of `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "mdcr") %>% pull(hr)` (`r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "mdcr") %>% pull(lower)` to `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "mdcr") %>% pull(upper)`; 95% CI),
`r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "optum_dod") %>% pull(hr)` (`r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "optum_dod") %>% pull(lower)` to `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "optum_dod") %>% pull(upper)`; 95% CI) and `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "panther") %>% pull(hr)` (`r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "panther") %>% pull(lower)` to `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "panther") %>% pull(upper)`; 95% CI) within the highest risk quarter of MDCR, OPTUM_DOD and PANTHER respectively. This translated in absolute risk reduction of `r hipFractureAbsolute %>% filter(database == "mdcr") %>% pull(estimate)`% (`r hipFractureAbsolute %>% filter(database == "mdcr") %>% pull(lower)`% to `r hipFractureAbsolute %>% filter(database == "mdcr") %>% pull(upper)`%; 95% CI), `r hipFractureAbsolute %>% filter(database == "optum_dod") %>% pull(estimate)`% (`r hipFractureAbsolute %>% filter(database == "optum_dod") %>% pull(lower)`% to `r hipFractureAbsolute %>% filter(database == "optum_dod") %>% pull(upper)`%; 95% CI), `r hipFractureAbsolute %>% filter(database == "panther") %>% pull(estimate)`% (`r hipFractureAbsolute %>% filter(database == "panther") %>% pull(lower)`% to `r hipFractureAbsolute %>% filter(database == "panther") %>% pull(upper)`%; 95% CI). The random effects meta-analytic calibrated hazard ratio within the highest risk quarter was `r metaCalibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101) %>% pull(hr)` (`r metaCalibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101) %>% pull(lower)` to `r metaCalibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101) %>% pull(upper)`; 95% CI) across the 3 databases.

@@

\newpage
# References
\setlength{\parindent}{-0.25in}
\setlength{\leftskip}{0.25in}
\noindent

<div id="refs"></div>

\setlength{\parindent}{0in}
\setlength{\leftskip}{0in}
\noindent

\newpage
```{r plotMeta, echo=FALSE, fig.cap="This is a caption", out.width = '100%'}
knitr::include_graphics(here::here("figures/plotMeta.tiff"))
```

>>>>

```{r, echo=FALSE, message=FALSE}
library(tidyverse)
library(knitr)
library(kableExtra)

@@

- **IBM MarketScan Medicare Supplemental Database (MDCR)** is a claims database representing retirees in the United States with primary or Medicare supplemental coverage
- **Optum De-Identified Clinformatics® Data Mart Database – Date of Death (Optum-DOD)** is a claims database including members of private health insurance.

@@

Our cohorts included female new users of teriparatide or oral bisphosphonates above the age of 65 [@Ryan2013]. Patients were considered to be new-users if they received no anti-osteoporosis drugs (raloxifene, bazedoxifene, denosumab, abaloparatide, romosozumab) 365 days prior to treatment initiation with teriparatide or an oral bisphosphonate. We required patients to have at least one year of observation period before first exposure and an established osteoporosis diagnosis, defined as any recorded fracture of the hip, wrist, spine or shoulder/humerus in their recorded medical history.

@@

##     Statistical analyses
We carried out two sets of analyses. First, we derived overall treatment effect estimates of teriparatide compared to oral bisphosphonates regarding the three outcomes of interest. To account for potential measured confounding we developed separate large-scale propensity score models within each database based on LASSO logistic regression using the same predefined set of measured covariates [@Tian2018]. We estimated treatment effects using Cox proportional hazards models with treatment as the sole covariate fitted in the 1:4 propensity score-matched subset of the considered patient population. Results from different databases were summarized using random effects meta-analysis.

For the second set of analyses, we used the standardized framework for risk-based assessment of treatment effect heterogeneity, that follows the directives of the PATH statement [@Kent2019; @KentEnE2019]. Following data extraction, the framework was applied in two steps. First, we derived individualized risk predictions for the three efficacy outcomes. We built the prediction models using LASSO logistic regression on the propensity score matched (1:4) subpopulation of the pooled treatment arms, aiming to remove any effect of measured confounding that may cause differential fit of our models across treatment arms. We considered the same large set of candidate covariates for the development of our prediction models.

For each outcome we used the derived prediction models to divide the population under study in 4 equally-sized subgroups, based on the quarters of the predicted risk distribution. Within each of these risk-based subgroups we developed a new propensity score model. Our analyses were performed on the propensity score matched (1:4) subset of the risk subgroup subset. We derived relative effect estimates using Cox proportional hazards models only with treatment as a predictor. Absolute effect estimates were calculated based on the difference of the Kaplan-Meier estimates, on day 730 after treatment initiation.

Residual study bias from unmeasured confounding can still be present in observational studies, which often may not be visible when evaluating propensity score adjustment. To account for that, we considered a set of 126 negative control outcome experiments, where a null effect was assumed to be true. We used the estimated relative effects for these outcomes to derive an empirical approximation to the true null distribution, which was then used to calibrate the hazard ratios and their 95% confidence intervals for the 3 outcomes of interest.

@@

When assessing heterogeneity of treatment effect for hip fracture, we estimated calibrated hazard ratios of `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "mdcr") %>% pull(hr)` (`r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "mdcr") %>% pull(lower)` to `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "mdcr") %>% pull(upper)`; 95% CI),
`r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "optum_dod") %>% pull(hr)` (`r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "optum_dod") %>% pull(lower)` to `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "optum_dod") %>% pull(upper)`; 95% CI) and `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "panther") %>% pull(hr)` (`r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "panther") %>% pull(lower)` to `r calibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101, database == "panther") %>% pull(upper)`; 95% CI) within the highest risk quarter of MDCR, Optum-DOD and Optum-EHR respectively. This translated in absolute risk reduction of `r hipFractureAbsolute %>% filter(database == "mdcr") %>% pull(estimate)`% (`r hipFractureAbsolute %>% filter(database == "mdcr") %>% pull(lower)`% to `r hipFractureAbsolute %>% filter(database == "mdcr") %>% pull(upper)`%; 95% CI), `r hipFractureAbsolute %>% filter(database == "optum_dod") %>% pull(estimate)`% (`r hipFractureAbsolute %>% filter(database == "optum_dod") %>% pull(lower)`% to `r hipFractureAbsolute %>% filter(database == "optum_dod") %>% pull(upper)`%; 95% CI), `r hipFractureAbsolute %>% filter(database == "panther") %>% pull(estimate)`% (`r hipFractureAbsolute %>% filter(database == "panther") %>% pull(lower)`% to `r hipFractureAbsolute %>% filter(database == "panther") %>% pull(upper)`%; 95% CI) respectively (Figure \@ref(fig:plotAbsoluteHip)). The random effects meta-analytic calibrated hazard ratio within the highest risk quarter was `r metaCalibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101) %>% pull(hr)` (`r metaCalibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101) %>% pull(lower)` to `r metaCalibrateRiskStratified %>% filter(riskStratum == "Q4", stratOutcome == 101, estOutcome == 101) %>% pull(upper)`; 95% CI) across the 3 databases. The full set of results can be explored at https://arekkas.shinyapps.io/ter_bis_3dbs/.

@@

\newpage
# References
\setlength{\parindent}{-0.25in}
\setlength{\leftskip}{0.25in}
\noindent
<div id="refs"></div>
\setlength{\parindent}{0in}
\setlength{\leftskip}{0in}
\noindent

\newpage
# Tables and figures
```{r, echo=FALSE}
incidenceOverall <- readRDS(here::here("data/raw/incidenceOverall.rds"))
incidence <- incidenceOverall %>% 
  filter(analysisType == "matchOnPs_1_to_4") %>%
  mutate(
    database = case_when(
      database == "mdcr" ~ "MDCR",
      database == "optum_dod" ~ "Optum-DOD",
      database == "panther" ~ "Optum-EHR"
    ),
    treatmentDays = round(treatmentDays / 365),
    comparatorDays = round(comparatorDays / 365)
  ) %>%
  select(
    database,
    outcomeId,
    treatmentPersons,
    treatmentDays,
    treatmentOutcomes,
    comparatorPersons,
    comparatorDays,
    comparatorOutcomes
  ) %>%
  arrange(database, outcomeId) %>%
  mutate(
    database = ifelse(
      outcomeId == 101,
      database,
      ""
    ),
    outcomeId = case_when(
      outcomeId == 101 ~ "Hip fracture",
      outcomeId == 102 ~ "Major osteoporotic\nfracture",
      outcomeId == 103 ~ "Vertebral fracture"
    ),
  )

kable(
  incidence,
  longtable = TRUE,
  booktabs = TRUE,
  col.names = c(
    "Database",
    "Outcome",
    "Persons",
    "Person years",
    "Events",
    "Persons",
    "Person years",
    "Events"
  ),
  format.args = list(
    big.mar = ","
  )
) %>%
  add_header_above(c("", "", "Teriparatide" = 3, "Bisphosphonates" = 3))
  # kableExtra::add_indent(c(1, 4, 7))
```

\newpage
```{r plotMeta, echo=FALSE, fig.cap="Calibrated hazard ratios for the 3 outcomes of interest across the 3 considered databases using a set of 126 negative controls. Values below 1 favour teriparatide, while values above 1 favour bisphosphonates.", out.width = '100%'}
knitr::include_graphics(here::here("figures/plotMeta.png"))
```

\newpage
```{r plotAbsoluteHip, echo=FALSE, fig.cap="Absolute risk differences wihtin quarters of predicted hip fracture risk across the 3 considered databases.", out.width = '100%'}
knitr::include_graphics(here::here("figures/plotAbsoluteHip.png"))
```
